---
title: "Getting started with y2conjoint"
vignette: >
  %\VignetteIndexEntry{Getting started with y2conjoint}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
library(y2conjoint)
library(dplyr)
library(ggplot2)
```

`y2conjoint` provides data structures and functions for working with the output
of a choice-based conjoint study. This article walks through a full example
using a simulated **cell-phone** study: loading raw utilities, describing the
attributes, defining products, estimating preference shares (overall and by
subgroup), and stress-testing a design with a sensitivity analysis.

The phones in this study vary on five attributes — **Brand**, **Price**,
**Battery** life, **Storage**, and **Color** — and the goal is to understand how
those attributes drive choice.

## The raw data

A conjoint model produces one *part-worth utility* per respondent per attribute
level. These arrive in a terse "model format" where columns are named
`A[NUM]B[NUM]` — attribute *i*, level *j*. The package ships a small simulated
example:

```{r}
#| label: raw-data
head(example_utilities)
```

Alongside the utilities we need a **crosswalk** that says what each column means:
which attribute (collection) it belongs to, its human-readable name, and — for
ordered attributes like price — its rank.

```{r}
#| label: crosswalk
example_crosswalk
```

Here `Brand` and `Color` are unordered (`collection_order` is `NA`), while
`Price`, `Battery`, and `Storage` are ordered.

## Building a conjoint_df

`conjoint_df()` renames the model-format columns to their user-facing names and
records the attribute structure as metadata. Extra columns (IDs, demographics)
are carried along untouched.

```{r}
#| label: build
cjt <- conjoint_df(example_utilities, example_crosswalk)
cjt
```

The print header summarizes the collections (ordered ones are annotated
`(ordered)`), the outside-good column, and how many extra columns are present.

### Protected columns

The level columns and the `NONE` outside good are **protected**: they hold the
utilities the whole analysis depends on, so `conjoint_df` guards them against
accidental loss.

```{r}
#| label: protected
protected_cols(cjt)
```

You can wrangle demographics freely, but a protected column cannot be dropped,
renamed, or overwritten by accident.

```{r}
#| label: filter-ok
# Filtering rows and editing demographics is fine.
cjt |>
  filter(age >= 40) |>
  mutate(age = age %/% 10 * 10)
```

```{r}
#| label: protected-error
#| error: true
# Overwriting a protected column is blocked.
cjt |>
  mutate(Northwind = 0)
```

If you genuinely need to reshape the utilities, call `tibble::as_tibble()` first
to drop the guardrails.

## Attributes are collections

`conjoint_df()` builds a [collection] for each attribute from the crosswalk, but
it is worth understanding them directly, because products (below) are validated
against them. A plain `collection()` describes an unordered attribute, while
`ordered_collection()` records a meaningful order — exactly the distinction
between a phone's brand and its price:

```{r}
#| label: collections
brand <- collection("Brand", c("Northwind", "Cascade", "Meridian"))
price <- ordered_collection("Price", c("$199", "$299", "$399"))

brand
price

is_ordered(brand)
is_ordered(price)
```

`is_ordered()` tells the two apart — useful when deciding whether it makes sense
to talk about "moving up a level" for an attribute.

### Absence levels

Some attributes include a level that means *having none of the feature*. If this
study had also tested a camera, "No camera" would be that level. Flag it with
`absence` so the rest of the package knows it is special:

```{r}
#| label: absence
camera <- collection(
  "Camera",
  c("No camera", "12 MP", "48 MP"),
  absence = "No camera"
)
camera
```

An absence level is mutually exclusive with other levels of the same attribute — a
phone cannot simultaneously have "No camera" and a 48 MP camera — and the package
enforces that when you build a product (next section).

## Describing products

A `product` describes one product by selecting levels across collections. Selecting
a single level per collection is the common case:

```{r}
#| label: product
flagship <- product(
  cjt,
  c("Meridian", "$399", "30 hours", "512 GB", "Black"),
  name = "Flagship"
)
flagship
```

You may select *several* levels within one collection — useful for a co-branded
phone, where the combine function (below) decides how to merge them:

```{r}
#| label: product-cobrand
co_brand <- product(
  cjt,
  c("Northwind", "Cascade", "Meridian", "$199", "10 hours", "128 GB", "Blue"),
  name = "Co_brand"
)

co_brand
```

Absence levels are the exception: because they mean "none of this feature", they
cannot be paired with another level from the same attribute. A product that tries
to is rejected:

```{r}
#| label: product-absence
#| error: true
product(
  list(brand, price, camera),
  c("Northwind", "$199", "No camera", "48 MP")
)
```

## From product to utility

`compute_product_utility()` is the building block underneath everything else: it
collapses a product into a single utility per respondent by combining the selected
levels within each attribute and summing across attributes.

```{r}
#| label: utility
head(compute_product_utility(cjt, flagship))
```

On its own this vector answers "how appealing is this phone to each respondent?".
Preference share (next) turns those numbers into a market forecast.

## Competitive sets and preference shares

A `competitive_set` bundles the phones that compete for choice. `run_scenario()`
applies a logit (softmax) choice model: it turns each respondent's product
utilities into choice probabilities, takes into account the NONE preference, and
averages the probabilities into mean preference shares.

```{r}
#| label: scenario
launch_market <- competitive_set(
  product(cjt, c("Northwind", "$199", "20 hours", "256 GB", "Black"), name = "Value"),
  product(cjt, c("Cascade", "$299", "20 hours", "256 GB", "Silver"), name = "Mainstream"),
  product(cjt, c("Meridian", "$399", "30 hours", "512 GB", "Black"), name = "Premium"),
  name = "Launch"
)

launch_market 

shares <- run_scenario(cjt, launch_market)
shares
```

The result is a plain tibble with one row per competitive set - here just
`Launch` - identified by its `competitive_set` column, and one `share_*` column
per phone plus `share_NONE` for respondents who would choose none of these
phones. Excluding `competitive_set`, the row sums to 1. `competitive_set()`'s
`none` argument (default `TRUE`) controls whether NONE competes at all; setting
`none = FALSE` forces a choice among the set's products and drops `share_NONE`
from the result.

```{r}
#| label: plot
#| fig-alt: "Bar chart of preference share by product."
#| fig-width: 6
#| fig-height: 3
shares |>
  tidyr::pivot_longer(
    starts_with("share_"),
    names_to = "product",
    values_to = "share"
  ) |>
  mutate(product = sub("^share_", "", product)) |>
  ggplot(aes(share, reorder(product, share))) +
  geom_col() +
  labs(x = "Preference share", y = NULL)
```

### Combining levels within a collection

When a product selects more than one level from a collection, `combine_fn`
decides how to merge them. The default, `pmax()`, takes the best available level
— a reasonable rule for co-branding, where buyers respond to the more appealing
brand. To instead model the two brands *reinforcing* one another, pass a **named
list** that sums the `Brand` utilities; any collection you omit falls back to
`pmax()`:

```{r}
#| label: combine-fn
co_market <- competitive_set(
  product(
    cjt,
    c("Northwind", "Cascade", "$199", "20 hours", "256 GB", "Black"),
    name = "Co-brand"
  ),
  product(cjt, c("Meridian", "$399", "30 hours", "512 GB", "Black"), name = "Premium"),
  name = "Co_brand launch"
)

run_scenario(cjt, co_market, combine_fn = list(Brand = `+`))
```

### Scaling choice probabilities

`scaling_factor` multiplies every utility before the softmax. Values above 1
sharpen the shares toward the most attractive phone; values below 1 flatten them
toward an even split. It is a convenient handle for calibrating a simulator to
known market conditions.

```{r}
#| label: scaling
run_scenario(cjt, launch_market, scaling_factor = 2)
```

## Preference shares by subgroup

Marketers rarely want a single number — they want to know *who* prefers *what*.
The `.by` argument computes shares within subgroups. Each column you name is
treated marginally: the sample is split by that column's values, scored within
each, and the results are stacked, with the subgroup size `n` reported alongside.

```{r}
#| label: by-region
run_scenario(cjt, launch_market, .by = region)
```

The `n` column matters: a share computed from a handful of respondents is noisy,
so always check the base before reading too much into a subgroup.

```{r}
#| label: by-education-plot
#| fig-alt: "Preference share by product across education groups."
#| fig-width: 7
#| fig-height: 3
run_scenario(cjt, launch_market, .by = education) |>
  tidyr::pivot_longer(
    starts_with("share_"),
    names_to = "product",
    values_to = "share"
  ) |>
  mutate(
    product = sub("^share_", "", product),
    group_level = factor(group_level, levels = unique(group_level))
  ) |>
  ggplot(aes(share, group_level)) +
  geom_col() +
  facet_wrap(vars(product)) +
  labs(x = "Preference share", y = NULL)
```

## Comparing several competitive sets {#comparing-several-competitive-sets}

`run_scenario()` also accepts a **list** of competitive sets. It scores each one
independently and binds the results together — the quickest way to compare whole
lineups side by side. Here we pit the launch lineup against a possible refresh:

```{r}
#| label: multi-set
refresh <- competitive_set(
  product(cjt, c("Cascade", "$199", "30 hours", "512 GB", "Blue"), name = "Value"),
  product(cjt, c("Meridian", "$299", "30 hours", "512 GB", "Black"), name = "Premium"),
  name = "Refresh"
)

comparison <- run_scenario(cjt, list(launch_market, refresh))
comparison
```

The result now has one row per competitive set (`Launch` and `Refresh`). Both
lineups sell a "Value" phone, so `share_Value` is shared by both rows - each
set's own share appears in its own row. "Mainstream" and "Premium" phones are
unique to one lineup, so `share_Mainstream`/`share_Premium` are `NA` in the
other set's row.

```{r}
#| label: multi-set-plot
#| fig-alt: "Bar chart of preference share for every product across both lineups."
#| fig-width: 6
#| fig-height: 3
comparison |>
  tidyr::pivot_longer(
    starts_with("share_"),
    names_to = "product",
    values_to = "share"
  ) |>
  mutate(product = sub("^share_", "", product)) |>
  ggplot(aes(share, reorder(product, share))) +
  geom_col() +
  facet_wrap(vars(competitive_set)) +
  labs(x = "Preference share", y = NULL)
```

## Sensitivity analysis

Once you have a candidate phone, the next question is usually "which change would
move share the most?". `sensitivity_analysis()` perturbs a product one level at a
time and reports how each change shifts preference share against a baseline.

For a single-select attribute it swaps the chosen level for each alternative
(`alone`); for an attribute you mark as multiple-select it instead adds
(`added_in`) or removes (`taken_out`) a level. Here we treat `Brand` as
multiple-select — modelling co-branding — and everything else as single-select:

```{r}
#| label: sensitivity
sens <- sensitivity_analysis(cjt, multiple_select = "Brand", product = flagship)
sens
```

A tornado chart makes the levers obvious: bars to the right raise share relative
to the flagship baseline, bars to the left cost share.

```{r}
#| label: sensitivity-plot
#| fig-alt: "Tornado chart of preference-share change for each perturbation."
#| fig-width: 7
#| fig-height: 4
sens |>
  mutate(change = paste0(Feature, ": ", Level, " (", comparison, ")")) |>
  ggplot(aes(delta, reorder(change, delta))) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  labs(
    x = "Change in preference share vs. flagship baseline",
    y = NULL
  )
```

Read this as a to-do list for the design: the changes at the extremes are where
price, battery, or storage decisions matter most for this phone's share.
```
