---
title: "Getting started with y2conjoint"
vignette: >
  %\VignetteIndexEntry{Getting started with y2conjoint}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
format: 
  html:
    code-overflow: wrap
    include-in-header:
      - text: |
          <style>
          /* Forces raw text execution output to wrap horizontally */
          div.cell-output pre code {
              white-space: pre-wrap !important;
              word-break: break-all !important;
          }
          </style>
---

```{r}
#| label: setup
#| include: false
library(y2conjoint)
library(dplyr)
library(ggplot2)
```

`y2conjoint` provides data structures and functions for working with the output
of a choice-based [conjoint study](https://online.hbs.edu/blog/post/what-is-conjoint-analysis). This article walks through a full example
using a simulated **cell-phone** study: loading raw utilities, describing the
attributes, defining products, estimating preference shares (overall and by
subgroup), and tweaking a design with a sensitivity analysis.

The phones in this study vary on five attributes — **Brand**, **Price**,
**Battery life**, **Storage**, and **Color** — and the goal is to understand how
those attributes drive choice.

## The raw data

A conjoint model produces one *part-worth utility* per respondent per attribute
level. These arrive in a terse "model format" where columns are named
`A[NUM-i]B[NUM-j]` — attribute *i*, level *j*. The package includes a small simulated
example:

```{r}
#| label: raw-data
head(example_utilities)
```

Alongside the utilities we need a **crosswalk** that says what each column means:
which attribute (collection) it belongs to, its human-readable name, and — for
ordered attributes like price — its rank. Creating the crosswalk is the thing that will take you the most time. 
It must be created manually. You need to carefully check the attributes and levels and make sure they are in the correct order. 

```{r}
#| label: crosswalk
example_crosswalk
```

Here `Brand` and `Color` are unordered (`collection_order` is `NA`), while
`Price`, `Battery`, and `Storage` are ordered.

A crosswalk must contain the following columns:
- old_name
- user_name
- collection_name

It may also include these additional columns:
- collection_order
- absense_levels

## Building a conjoint_df

`conjoint_df()` takes the utilities (model output), and the crosswalk, and returns a conjoint dataframe - the basis for all analysis. It renames the model-format columns to their user-facing names and records the attribute structure as metadata. 
Extra columns (IDs, demographics) are carried along untouched.

```{r}
#| label: build
cjt <- conjoint_df(example_utilities, example_crosswalk)
cjt
```

You may print a conjoint dataframe to get useful meta-data. 
The print header summarizes the collections (ordered ones are annotated
`(ordered)`), which column was modeled as the NONE column, and how many extra (non-utility) columns are present.

### Protected columns

The level columns and the `NONE` outside good are **protected**: they hold the
utilities the whole analysis depends on, so `conjoint_df` guards them against
accidental loss.

```{r}
#| label: protected
protected_cols(cjt)
```

You can wrangle demographics freely, but a protected column cannot be dropped,
renamed, or overwritten by accident.

```{r}
#| label: filter-ok
# Filtering rows and editing demographics is fine.
cjt |>
  filter(age >= 40) |>
  mutate(age = age %/% 10 * 10)
```

```{r}
#| label: protected-error
#| error: true
# Overwriting a protected column is blocked.
cjt |>
  mutate(Northwind = 0)
```

If you genuinely need to reshape the utilities, call `tibble::as_tibble()` to convert the conjoint 
dataframe into a tibble which drops the guardrails.

## Attributes are collections

`conjoint_df()` builds a `collection()` for each attribute from the crosswalk, but
it is worth understanding them directly, because products (below) are validated
against them. A plain `collection()` describes an unordered attribute, while
`ordered_collection()` records a meaningful order — exactly the distinction
between a phone's brand and its price. `collection()` takes the name of the attribute/collection, 
and a list of levels. 

```{r}
#| label: collections
brand <- collection(
  name = "Brand",
  levels = c("Northwind", "Cascade", "Meridian")
)
price <- ordered_collection(name = "Price", levels = c("$199", "$299", "$399"))

brand
price

is_ordered(brand)
is_ordered(price)
```

`is_ordered()` tells the two apart — useful when deciding whether it makes sense
to talk about "moving up a level" for an attribute.

### Absence levels

Some attributes include a level that means *having none of the feature*. If this
study had also tested a camera, "No camera" would be that level. Flag it with
`absence` so the rest of the package knows it is special:

```{r}
#| label: absence
camera <- collection(
  name = "Camera",
  levels = c("No camera", "12 MP", "48 MP"),
  absence = "No camera"
)

camera
```

An absence level is mutually exclusive with other levels of the same attribute — a
phone cannot simultaneously have "No camera" and a 48 MP camera — and the package
enforces that when you build a product (next section).

## Describing products

A `product` describes one product by selecting levels across collections. Selecting
a single level per collection is the common case:

```{r}
#| label: product
flagship <- product(
  x = cjt,
  levels = c("Meridian", "$399", "30 hours", "512 GB", "Black"),
  name = "Flagship"
)
flagship
```

You may select *several* levels within one collection — useful for a co-branded
phone, where the combine function (below) decides how to merge them. You do not have to specify which 
collection a level is a part of since it is required that levels are unique.

```{r}
#| label: product-cobrand
co_brand <- product(
  x = cjt,
  levels = c(
    "Northwind",
    "Cascade",
    "Meridian",
    "$199",
    "10 hours",
    "128 GB",
    "Blue"
  ),
  name = "Co_brand"
)

co_brand
```

Absence levels are the exception: because they mean "none of this feature", they
cannot be paired with another level from the same attribute. A product that tries
to is rejected:

```{r}
#| label: product-absence
#| error: true
product(
  x = list(brand, price, camera),
  levels = c("Northwind", "$199", "No camera", "48 MP")
)
```

Products can take a `conjoint_df()` object, or a list of collections to validate against. You may want to pass lists of collections if you do bridging work across multiple studies. 

## From product to utility

`compute_product_utility()` is the building block underneath everything else: it
collapses a product into a single utility per respondent by combining the selected
levels within each attribute and summing across attributes.

```{r}
#| label: utility
compute_product_utility(x = cjt, product = flagship) |>
  head()
```

On its own this vector answers "how appealing is this phone to each respondent?".
Preference share (next) turns those numbers into a market forecast.

## Competitive sets and preference shares

A `competitive_set` bundles the phones that compete for choice. `run_scenario()`
applies a logit (softmax) choice model: it turns each respondent's product
utilities into choice probabilities, takes into account the NONE preference, and
averages the probabilities into mean preference shares.

```{r}
#| label: scenario
value <- product(
  cjt,
  c("Northwind", "$199", "20 hours", "256 GB", "Black"),
  name = "Value"
)
mainstream <- product(
  cjt,
  c("Cascade", "$299", "20 hours", "256 GB", "Silver"),
  name = "Mainstream"
)
premium <- product(
  cjt,
  c("Meridian", "$399", "30 hours", "512 GB", "Black"),
  name = "Premium"
)

launch_market <- competitive_set(
  value,
  mainstream,
  premium,
  name = "Launch"
)

launch_market

shares <- run_scenario(cjt, launch_market)
shares
```

The result is a plain tibble with one row per competitive set - here just
`Launch` - identified by its `competitive_set` column, and one `share_*` column
per phone plus `share_NONE` for respondents who would choose none of these
phones. Excluding `competitive_set`, the row sums to 1. `competitive_set()`'s
`none` argument (default `TRUE`) controls whether NONE competes at all; setting
`none = FALSE` forces a choice among the set's products and drops `share_NONE`
from the result.

You can easily visualize the outputs from run_scenario and plot them:

```{r}
#| label: plot
#| fig-alt: "Bar chart of preference share by product."
#| fig-width: 6
#| fig-height: 3
shares |>
  tidyr::pivot_longer(
    starts_with("share_"),
    names_to = "product",
    values_to = "share"
  ) |>
  mutate(product = sub("^share_", "", product)) |>
  ggplot(aes(share, reorder(product, share))) +
  geom_col() +
  labs(x = "Preference share", y = NULL)
```

### Combining levels within a collection

When a product selects more than one level from a collection, `combine_fn`
decides how to merge them. The default, `pmax()`, takes the best available level
— a reasonable rule for co-branding, where buyers respond to the more appealing
brand. To instead model the two brands *reinforcing* one another, pass a 
**named list** that sums the `Brand` utilities; any collection you omit falls 
back to `pmax()`. Combination functions can be set per-collection, not per-product. 
This ensures that we are comparing products with the same market assumptions.

```{r}
#| label: combine-fn

co_brand <- product(
  x = cjt,
  levels = c("Northwind", "Cascade", "$199", "20 hours", "256 GB", "Black"),
  name = "Co-brand"
)

co_market <- competitive_set(
  co_brand,
  premium,
  name = "Co_brand launch"
)

run_scenario(
  x = cjt,
  competitive_set = co_market,
  combine_fn = `+`
) # Sets the combination function as + for all collections

run_scenario(
  cjt,
  co_market,
  combine_fn = list(Brand = `+`)
) # Sets the combination function as + for just Brand
```

### Scaling choice probabilities

`scaling_factor` multiplies every utility before the softmax. Values above 1
sharpen the shares toward the most attractive phone; values below 1 flatten them
toward an even split. It is a convenient handle for calibrating a simulator to
known market conditions. If you know what sales for a certain product are relative 
to another product, you can increase or decrease the scaling factor until the 
simulator matches the known market conditions. 

```{r}
#| label: scaling
run_scenario(cjt, launch_market, scaling_factor = 2)
```

## Preference shares by subgroup

Marketers rarely want a single number — they want to know *who* prefers *what*.
The `.by` argument computes shares within subgroups. Each column you name is
treated marginally: the sample is split by that column's values, scored within
each, and the results are stacked, with the subgroup size `n` reported alongside.

```{r}
#| label: by-region
run_scenario(cjt, launch_market, .by = region)
```

The `n` column matters: a share computed from a handful of respondents is noisy,
so always check the base before reading too much into a subgroup.

Again, the output of run_scenario is easy to plot:

```{r}
#| label: by-education-plot
#| fig-alt: "Preference share by product across education groups."
#| fig-width: 7
#| fig-height: 3
run_scenario(cjt, launch_market, .by = education) |>
  tidyr::pivot_longer(
    starts_with("share_"),
    names_to = "product",
    values_to = "share"
  ) |>
  mutate(
    product = sub("^share_", "", product),
    group_level = factor(group_level, levels = unique(group_level))
  ) |>
  ggplot(aes(share, group_level)) +
  geom_col() +
  facet_wrap(vars(product)) +
  labs(x = "Preference share", y = NULL)
```

## Comparing several competitive sets {#comparing-several-competitive-sets}

`run_scenario()` also accepts a **list** of competitive sets. It scores each one
independently and binds the results together — the quickest way to compare whole
lineups side by side. Here we pit the launch lineup against a market where we 
discontinue the mainstream product:

```{r}
#| label: multi-set
deprecate_mainstream <- competitive_set(
  value,
  premium,
  name = "Deprecate Mainstream"
)

comparison <- run_scenario(cjt, list(launch_market, deprecate_mainstream))
comparison
```

The result now has one row per competitive set (`Launch` and `deprecate_mainstream`). Both
lineups sell a "Value" phone and a "Premium" phone, so `share_Value` and `share_Premium` are 
shared by both rows - each set's own share appears in its own row. "Mainstream"  phones are
unique to the launch_market lineup, so `share_Mainstream` are `NA` in the deprecate_mainstream row.

```{r}
#| label: multi-set-plot
#| fig-alt: "Bar chart of preference share for every product across both lineups."
#| fig-width: 6
#| fig-height: 3
comparison |>
  tidyr::pivot_longer(
    starts_with("share_"),
    names_to = "product",
    values_to = "share"
  ) |>
  mutate(product = sub("^share_", "", product)) |>
  ggplot(aes(share, reorder(product, share))) +
  geom_col() +
  facet_wrap(vars(competitive_set)) +
  labs(x = "Preference share", y = NULL)
```

## Sensitivity analysis

Once you have a candidate phone, the next question is usually "which change would
move share the most?". `sensitivity_analysis()` modifies a product one level at a
time and reports how each change shifts preference share against a baseline.

For a single-select attribute it swaps the chosen level for each alternative
(`alone`); for an attribute you mark as multiple-select it instead adds
(`added_in`) or removes (`taken_out`) a level. Here we treat `Brand` as
multiple-select — modelling co-branding — and everything else as single-select:

```{r}
#| label: sensitivity
sens <- sensitivity_analysis(cjt, multiple_select = "Brand", product = flagship)
sens
```

A tornado chart makes the levers obvious: bars to the right raise share relative
to the flagship baseline, bars to the left cost share.

```{r}
#| label: sensitivity-plot
#| fig-alt: "Tornado chart of preference-share change for each perturbation."
#| fig-width: 7
#| fig-height: 4
sens |>
  mutate(change = paste0(Feature, ": ", Level, " (", comparison, ")")) |>
  ggplot(aes(delta, reorder(change, delta))) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  labs(
    x = "Change in preference share vs. flagship baseline",
    y = NULL
  )
```

Read this as a to-do list for the design: the changes at the extremes are where
price, battery, or storage decisions matter most for this phone's share.

